---
title: "Mice Protein Expression"
subtitle: |
  Statistical Learning |
  Master in Statistics for Data Science
author: 
- "Amara Scheel"
- "Noon Elsheikh"
- "Alberto Peces Contreras"
date: "2026-21-02"
output: 
  pdf_document:
    number_sections: true
    toc: true
geometry: margin=2cm
header-includes:
  - \usepackage{float}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.pos = 'H')
```

\newpage

# Introduction

## Data description

**Research Context:** This project is based on the research article
*"Self-Organizing Feature Maps Identify Proteins Critical to Learning in
a Mouse Model of Down Syndrome."*

**Dataset Source:** [UCI Machine Learning Repository: Mice Protein
Expression](https://doi.org/10.24432/C50S3Z)

**Biological Background:** Down syndrome (DS) is a chromosomal
abnormality (trisomy of human chromosome 21) associated with
intellectual disability. The overexpression of genes encoded by the
extra chromosome is believed to perturb standard pathways, causing
learning deficits.

**Goal:** We design a strategy based on probabilistic
classifiers—Gaussian Naive Bayes (GNB), Linear Discriminant Analysis
(LDA), and Quadratic Discriminant Analysis (QDA)—to identify
biologically important differences in protein expression levels in mice.

We include Logistic Regression as a discriminative baseline model. While
LDA/QDA model the joint distribution of the predictors (generative),
Logistic Regression directly models the posterior probabilities,
providing a robust, assumption-light benchmark against which to evaluate
our Gaussian models.

**Structure:** The dataset comprises expression levels of 77 proteins
measured across 1007 tissue samples. The samples are divided into two
main Genotypes:

-   **Control:** Normal genotype mice.

-   **Ts65Dn:** Trisomic (Down Syndrome) mice.

Additional factors include Treatment (Saline vs. Memantine) and Behavior
(Stimulated vs. Not Stimulated). For this midterm project, our primary
target variable is the **Genotype** (*Control* vs. *Ts65Dn*).

# Data Preparation

## Auxiliary tools

We define helper functions to automate repetitive tasks: `checkNA` for
visualizing missing data, `imputation` for filling missing values, and
`SelectKBest` for filtering features using ANOVA.

-   `checkNA`: It was built for analyzing and visualizing missing data.
    It calculates the proportion of missing values for every column in
    the dataset. It generates a bar plot showing the proportion of
    missingness per feature.

-   `SelectKBest`: It performs feature selection to find the most
    relevant variables. It runs an ANOVA test for every feature `x`
    against a target variable `y`.

-   `get_group_medians`: It calculates specific, subgroup-level medians.
    Instead of calculating a single global median for a column, it
    groups the data by `Genotype` and `Treatment`.

-   `impute_with_medians`: It performs imputation using subgroup
    medians. It looks for missing values in the specified
    `proteins`columns, and replaces them.

```{r auxiliary_tools, echo=FALSE}
checkNA <- function(data){
  # For every column, we count the number of NA
  na_counts = colSums(is.na(data))
  
  # Default return if no NAs
  if(sum(na_counts) == 0) {
    return(character(0))
  }

  # Compute (Nº columns with NA) / (Nº columns)
  if (sum(na_counts > 0) > 0){
    p = round(length(names(na_counts[na_counts > 0])) / ncol(data), 2)
    na_prop = sort(na_counts[na_counts > 0] / nrow(data), decreasing = TRUE)
  } else {
    p = 0
    na_prop = numeric(0)
  }

  # BARPLOT (Only plot if there are actually NAs to show)
  if(length(na_prop) > 0) {
      barplot(na_prop,las = 2, cex.names = 0.6, ylab = "Proportion of NA values",
              main = "Missing-value proportion across features", col = "gray")
      abline(h = 0.05, col = "red", lty = 2)
      legend("topright", 
             legend = c(paste0("Features with NA values: ", p*100, "%"),
                                              "Threshold for high missingness"),
             col = c("gray","red"), pch = c(15, NA), lty = c(NA, 2), bty = "n")
      grid()
  }
  
  # Return names of columns with > 5% missing data
  return (names(na_prop[na_prop > 0.05]))
}

SelectKBest <- function(X, y, K = 10){
  # Define auxiliary parameters
  n_features = ncol(X)
  f_score = numeric(n_features)
  names(f_score) = colnames(X)
  
  # For each feature, we compute ANOVA test between groups
  for (j in 1:n_features){
    # Safe extraction of F-value to prevent crash
    tryCatch({
      fit <- aov(X[,j] ~ y)
      summ <- summary(fit)
      if(!is.null(summ[[1]][["F value"]])) {
         f_score[j] = summ[[1]][["F value"]][1]
      } else {
         f_score[j] = 0 
      }
    }, error = function(e) { f_score[j] = 0 })
  }
  # We sort the F scores in decreasing order and select the best K features
  K = min(K, n_features)
  top_K = names(sort(f_score, decreasing = TRUE))[1:K]
  return (top_K)
}

# Compute subgroup medians from sepecific set and columns
get_group_medians <- function(df, proteins) {
  df %>%
    group_by(Genotype, Treatment) %>%
    summarise(across(all_of(proteins),
              \(x) median(x, na.rm = TRUE)), .groups = "drop")
}

# Impute using pre-computed medians
impute_with_medians <- function(df, medians, proteins) {
  for (col in proteins) {
    na_idx <- which(is.na(df[[col]]))
    for (i in na_idx) {
      ref_val <- medians %>%
        filter(Genotype  == df$Genotype[i],
               Treatment == df$Treatment[i]) %>%
        pull(col)
      df[[col]][i] <- ref_val
    }
  }
  return(df)
}
```

## Packages

```{r packages}
library(tidyverse)
library(knitr) # For tables
library(ggplot2)
library(patchwork)
library(caret)
library(class)
library(pROC)
library(e1071) # For Naive Bayes
library(MASS)  # For LDA/QDA
library(klaR)  # Additional tools
```

## Data Loading & Imputation

The first step is to identify missing values. When handling missing
data, it is important to consider the underlying mechanism: Missing
Completely at Random (MCAR), Missing at Random (MAR), or Missing Not at
Random (MNAR).

In proteomic datasets, high rates of missingness are frequently MNAR,
occurring when protein expression is too low for detection. Applying
median imputation to MNAR data would artificially inflate the unobserved
low values up to the group median, creating systemic bias.

To mitigate this risk, we adopt a threshold:

-   **Removal (\>5% missingness):** Features with a high proportion of
    missing data are assumed to carry a high risk of being MNAR and are
    excluded entirely from subsequent analyses to prevent
    imputation-induced bias.

-   **Imputation (\<5% missingness):** For variables with minimal
    missingness, we assume the absence is largely due to technical noise
    or measurement error (MCAR/MAR). To preserve these features without
    dropping valuable patient rows, we apply median imputation.

```{r data prep, echo = FALSE, fig.width=7, fig.height=4, width.out="100%", fig.align="center"}
# ---------
# Load data
# ---------
#!!! seperator may be ; or , depending on the dataset !!!
data = read.csv("Data_Cortex_Nuclear.csv", sep = ",", header = TRUE)
colNames = colnames(data)
# ----------------
# Different labels
# ----------------
metadata = c("MouseID", "Genotype", "Treatment", "Behavior", "class")
# ----------------------------------------
# Missing-value proportion across features
# ----------------------------------------
cols_with_NA = checkNA(data)
proteins = colNames[!(colNames %in% c(metadata, cols_with_NA))]
```

**Imputation Strategy:** To prevent data leakage, we first perform a
stratified 70/30 train/test split before any imputation takes place. We
then divide the training set into 4 biological subgroups (Control/Ts65Dn
× Memantine/Saline) and compute the median protein expression within
each group exclusively from the training data. These medians are
subsequently applied to impute missing values in both the training and
test sets.

This ensures two things: first, that the characteristics of treated mice
do not artificially influence control mice imputation values; and
second, that the test set is imputed using only information available at
training time, faithfully simulating a real deployment scenario where
new observations arrive without known labels.

```{r mean value imputation, echo = FALSE}
set.seed(619)
# --------------------------
# 1. Split BEFORE imputation
# --------------------------
lvls    <- c("Control", "Ts65Dn")
classes <- factor(data$Genotype, levels = lvls)

fltr_cols  <- !(colNames %in% cols_with_NA)
data_clean <- data[, fltr_cols]

idx_train  <- createDataPartition(classes, p = 0.7, list = FALSE)
data_train <- data_clean[idx_train, ]
data_test  <- data_clean[-idx_train, ]

# -------------------------------------------------------------------------
# 2. Compute subgroup medians from TRAINING SET only and perform imputation
# -------------------------------------------------------------------------
train_medians <- get_group_medians(data_train, proteins)
data_train <- impute_with_medians(data_train, train_medians, proteins)
data_test  <- impute_with_medians(data_test,  train_medians, proteins)

# ---------------------------------------------------------------------------
# 3. Reconstruct data_filtered to preserve compatibility with downstream code
# ---------------------------------------------------------------------------
data_filtered <- rbind(data_train, data_test)

# -----------------------
# 4. Verify no NAs remain and extract X_train/X_test
# -----------------------
cols_for_removal <- checkNA(data_filtered)
X_train <- data_train[, proteins]
X_test  <- data_test[, proteins]
```

## Multicollinearity

LDA and QDA require invertible covariance matrices; high
multicollinearity renders these singular or unstable. We empirically
determined that thresholds above 0.70 caused QDA to fail due to
ill-conditioned class covariance matrices, so we fix our cutoff at 0.70.
Features are removed using `findCorrelation` from `caret`, which
iteratively drops only the single most correlated variable from each
pair, preserving maximum predictive information.

```{r, multicollinearity handling, echo = FALSE, fig.width=6, fig.height=3.5, width.out="100%", fig.align="center"}
# --------------------------
# Compute correlation matrix
# --------------------------
cor_mat <- cor(data_filtered[,proteins])
cor_df <- as.data.frame(as.table(cor_mat))
colnames(cor_df) <- c("X1", "X2", "Correlation")
# ------------
# Heatmap plot
# ------------
ggplot(cor_df, aes(x = X1, y = X2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  ggtitle("Heatmap of Correlation across Features")
```

```{r correlation, echo = FALSE, fig.width=5.5, fig.height=4, width.out="100%", fig.align="center"}
# Identify correlated features to remove (>0.7)

# We use caret::findCorrelation which removes only ONE 
# feature from a highly correlated pair (the one with the highest 
# average correlation to all other features).

cut_off = round(seq(0.0, 1.0, length.out = 25), 2)
num_features = numeric(length(cut_off))

# Visualize impact of thresholds
for (i in seq_along(cut_off)){
  # Find how many features would be removed at each cutoff
  to_remove = findCorrelation(cor_mat, cutoff = cut_off[i])
  num_features[i] = length(to_remove)
}

# Apply the chosen 0.70 threshold
high_cor_idx = findCorrelation(cor_mat, cutoff = 0.7, names = TRUE)
n_high_cor = length(high_cor_idx)


# Select uncorrelated proteins
# Remove the identified highly correlated features from our protein list
if(length(high_cor_idx) > 0) {
    proteins = colNames[!(colNames %in% c(metadata, cols_with_NA, high_cor_idx))]
} else {
    proteins = colNames[!(colNames %in% c(metadata, cols_with_NA))]
}

# Plot: Number of variables removed vs. cut-off
plot(cut_off, num_features, type = "b", pch = 19, col = "blue",
     xlab = "Correlation Cut-off", ylab = "Number of features removed",
     main = "Features Removed vs. Correlation Cut-off")
abline(v = 0.7, col = "red", lty = 2)
grid()
legend("topright", legend = paste("Cutoff = 0.7\nRemoved =", n_high_cor), 
       col = "red", lty = 2, cex = 0.8, bty = "n")
```

# Exploratory Data Analysis (EDA)

Before modeling, we must thoroughly investigate the dataset's target
distribution, the underlying shapes of the feature distributions (as our
chosen probabilistic models heavily rely on normality assumptions), and
the multivariate expression profiles across classes.

## Class Imbalance Assessment

First, we examine the distribution of our target variable, `Genotype`.
Significant class imbalance can skew model training and render standard
accuracy metrics misleading.

```{r class_balance, echo = FALSE, fig.height=2.5, fig.width=5, fig.align="center"}
# Calculate proportions
class_counts <- table(data_filtered$Genotype)
class_props <- prop.table(class_counts)

ggplot(data_filtered, aes(x = Genotype, fill = Genotype)) +
  geom_bar(alpha = 0.8, color = "black", width = 0.5) + 
  geom_text(stat = "count", aes(label = ..count..), hjust = -0.1) +
  theme_minimal() +
  labs(title = "Target Variable Distribution: Genotype",
       y = "Count", x = "") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme(legend.position = "none") +
  coord_flip()

class_props_df <- data.frame(
  Genotype = names(class_props),
  Proportion = round(as.numeric(class_props), 3)
)

kable(class_props_df, caption = "Class distribution")
```

**Interpretation:** The bars for Control and Ts65Dn are roughly the same
height. This means Accuracy is a reliable metric, and the model has
enough examples of both classes to learn fairly.

## Feature Distribution and Normality Assumption

GNB, LDA, and QDA inherently assume that the continuous features follow
a Gaussian (Normal) distribution within each class. To validate this
assumption, we identify the single most discriminative protein using an
ANOVA F-test and examine its distribution via Density and Q-Q
(Quantile-Quantile) plots.

```{r normality, echo = FALSE, fig.height=3.25, fig.width=7, fig.align="center"}

# 1. Use SelectKBest to find the #1 best feature
top_feature_name <- SelectKBest(data_filtered[, proteins], data_filtered$Genotype, K = 1)

# 2. Density Plot
p1 <- ggplot(data_filtered, aes(x = .data[[top_feature_name]], fill = Genotype)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = paste("Density of", top_feature_name),
       x = "Expression Level", y = "Density") +
  scale_fill_brewer(palette = "Set1")

# 3. Q-Q Plot
p2 <- ggplot(data_filtered, aes(sample = .data[[top_feature_name]], color = Genotype)) +
  stat_qq(alpha = 0.5) + stat_qq_line(color = "black", linetype = "dashed") +
  theme_minimal() +
  labs(title = paste("Q-Q Plot of", top_feature_name),
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  scale_color_brewer(palette = "Set1")

# Display side by side
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Interpretation:** The density plot shows a clear separation between
the group means, confirming its predictive power. The Q-Q plot shows
that the quantiles of the data generally follow the theoretical diagonal
line of a normal distribution. While there are slight deviations in the
extreme tails, the overall Gaussian assumption holds sufficiently well
to proceed with LDA and QDA.

## Inter-group Protein Expression

To move beyond univariate analysis, we extract the top 10 most
discriminative proteins and visualize their average expression profiles
across the healthy and trisomic classes.

```{r topten, echo = FALSE, fig.height=4, fig.width=7, fig.align="center"}
# Get top 10 features
top_10_features <- SelectKBest(data_filtered[, proteins], data_filtered$Genotype, K = 10)

# Compute mean expression of these top 10 proteins per Genotype
heatmap_data <- data_filtered %>%
  group_by(Genotype) %>%
  summarise(across(all_of(top_10_features), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = -Genotype, names_to = "Protein", values_to = "Mean_Expression")

# Plot Heatmap
ggplot(heatmap_data, aes(x = Protein, y = Genotype, fill = Mean_Expression)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = round(Mean_Expression, 3)),  # <-- prints values on tiles
  color = "white", size = 3.5, fontface = "bold") +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 12, face = "bold")) +
  labs(title = "Expression Profiles of Top 10 Predictor Proteins",
       subtitle = "Average expression levels grouped by Genotype",
       x = "Protein", y = "", fill = "Mean Expr.")


```

```{r effectsizes, echo = FALSE, fig.height=3.5, fig.width=5, fig.align="center"}
effect_sizes <- data_filtered %>%
  summarise(across(all_of(top_10_features), ~ {
    ctrl  <- .x[data_filtered$Genotype == "Control"]
    ts    <- .x[data_filtered$Genotype == "Ts65Dn"]
    pooled_sd <- sqrt((var(ctrl) + var(ts)) / 2)
    (mean(ts) - mean(ctrl)) / pooled_sd   # Cohen's d
  })) %>%
  pivot_longer(everything(), names_to = "Protein", values_to = "Cohens_d")

# cat("Effect Sizes (Cohen's d) per Protein\n")
# print(effect_sizes %>% arrange(desc(abs(Cohens_d))), digits = 3)

ggplot(effect_sizes, aes(x = reorder(Protein, abs(Cohens_d)), 
                          y = Cohens_d, 
                          fill = Cohens_d > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("Higher in Control", "Higher in Ts65Dn")) +
  theme_minimal() +
  labs(title = "Effect Size (Cohen's d) for Top 10 Proteins",
       subtitle = "Captures meaningful difference relative to variability",
       x = "Protein", y = "Cohen's d", fill = "Direction")

kable(effect_sizes, digits = 3, caption = "Effect Sizes (Cohen's d) per Protein")
```

**Interpretation:** Among the top 10 proteins, raw mean expression
differences between genotypes are small (typically 0.05–0.10), which is
expected given that protein expression data is already normalized —
small absolute differences do not mean the proteins are unimportant.

Effect size analysis shows that two proteins have large effects
($d > 0.8$, one above $d = 1.0$), meaning the difference between groups
is larger than the natural variation within groups which is a strong
signal. Four proteins show medium effects ($d = 0.5$), and the remaining
three are weaker at $d = 0.3$. Overall, classification appears to be
driven by a small set of highly discriminant proteins, with the rest
providing supporting information.

## PCA

Finally, we perform Principal Component Analysis (PCA) on the test set to evaluate global class separability in lower dimensions.

```{r PCA, echo = FALSE, fig.height=4.5, fig.width=6, fig.align="center"}
# Compute PCA on the protein data
pca_res <- prcomp(data_train[, proteins], scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_df$Class <- data_train$Genotype

ggplot(pca_df, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(level = 0.95, linetype = 2) + 
  theme_minimal() +
  labs(title = "PCA: First Two Principal Components",
       subtitle = "Visualizing Class Separation",
       x = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100, 1), "%)")) +
  scale_color_brewer(palette = "Set1")
```

**Interpretation of PCA:** The first two principal components explain
approximately 36% of the total variance. The visualization shows a
significant overlap between the *Control* and *Ts65Dn* classes,
indicating that they are not linearly separable in these lower
dimensions.

```{r scree plot, echo = FALSE, fig.height=4, fig.width=6, fig.align="center"}
# Cumulative variance explained
var_explained <- summary(pca_res)$importance[3, ]  # cumulative proportion
n_components <- which(var_explained >= 0.90)[1]    # components for 90%

plot(var_explained, type = "b", pch = 19, col = "blue",
     xlab = "Number of Principal Components",
     ylab = "Cumulative Variance Explained",
     main = "Scree Plot: Cumulative Variance Explained")
abline(h = 0.90, col = "red", lty = 2)
abline(v = n_components, col = "red", lty = 2)
grid()

n_80 <- which(var_explained >= 0.80)[1]    # components for 80%
n_90 <- which(var_explained >= 0.90)[1]    # components for 90%

# Print statements
cat(sprintf("80%% of variance is explained by %d PCs\n", n_80))
cat(sprintf("90%% of variance is explained by %d PCs\n", n_90))
```

**Interpretation of the Scree Plot:** The dataset is highly
multi-dimensional, to explain 80% to 90% of the variance we require 10
to 17 principal components, respectively. This explains the overlap in
the 2D PCA plot and confirms that the classes are not linearly separable
in low dimensions. It justifies our use of multivariate probabilistic
models (LDA, QDA).

# Modeling Strategy

## Auxiliary tools for classifiers

We define some functions for each classifier, improving the future
pipeline structure. They all follow a similar pattern: train a model on
a training set, make prediction on a test set, and compute accuracy as
score.

-   `LogReg_classifier`: It builds a logistic regression model using the
    `glm()`function to predict a binary target variable `y`. If the
    probability is greater than 0.5, it assigns the class Ts65Dn;
    otherwise, it assigns Control.

-   `GNB_classifier`: It trains a Naive Bayes model, which calculates
    the probability of each class assuming independence and gaussianity.

-   `LDA_classifier`: It builds an LDA model, where a weighted
    covariance matrix is used for both classes.

-   `QDA_classifier`: It trains a QDA model, where each class have its
    own covariance matrix.

```{r classifiers, echo = FALSE}
# -------------------
# Logistic Regression
# -------------------
LogReg_classifier <- function(df_train, df_test, y_test){
  
  # Build the LogReg model log(mu) = beta_0 + X1*beta_1 + ... + Xn*beta_n
  model = glm(y ~ ., data = df_train, family = binomial(link = "logit"))
  
  # Compute the probability corresponding to each observation
  probs = predict(model, newdata = df_test, type = "response")
  
  # Calculate the classes manually
  y_pred = factor(ifelse(probs > 0.5, "Ts65Dn", "Control"), 
                  levels = c("Control", "Ts65Dn"))
  # Compute the accuracy
  acc = mean(y_pred == y_test)
  return(acc)
}
# --------------------
# Gaussian Naive Bayes
# --------------------
GNB_classifier <- function(df_train, df_test, y_test){

  # Build the Gaussian Naive Bayes model
  model = naiveBayes(y ~ ., data = df_train)
  
  # Calculate the predicted classes
  y_pred = predict(model, df_test, type = "class")
  
  # Compute the accuracy
  acc = mean(y_pred == y_test)
  return(acc)
}
# ----------------------------
# Linear Discriminant Analysis
# ----------------------------
LDA_classifier <- function(df_train, df_test, y_test){
  
  # Build the Linear Discriminant Analysis model
  model = lda(y ~ ., data = df_train)
  
  # Calculate the predicted classes
  pred = predict(model, df_test)
  y_pred = pred$class
  
  # Compute the accuracy
  acc = mean(y_pred == y_test)
  return(acc)
}
# -------------------------------
# Quadratic Discriminant Analysis
# -------------------------------
QDA_classifier <- function(df_train, df_test, y_test){
  tryCatch({
    model = qda(y ~ ., data = df_train)
    pred  = predict(model, df_test)
    y_pred = pred$class
    acc = mean(y_pred == y_test)
    return(acc)
  }, error = function(e){
    message("QDA failed (rank deficiency): returning NA")
    return(NA)
  })
}
```

## Feature Selection Pipeline

The pipeline begins by preprocessing and feature selection to prepare
the data for classification.

**Z-score scaling**:

-   From the training dataset, the mean $(\mu)$ and standard deviation
    $(\sigma)$ of each feature are computed.

-   Both the training and testing sets are then standardized using these
    parameters.

-   This ensures that all features contribute equally to the
    classifiers, avoiding bias from features with larger magnitudes.

**SelectKBest** with ANOVA:

-   The pipeline selects the **top K features** that have the strongest
    univariate relationship with the target class, based on the ANOVA
    F-score.

-   The value of `K` is defined before training and will be varied to
    determine the optimal number of features for each classifier.

**Classification**:

-   Each classifier (LogReg, GNB, LDA, and QDA) is trained using only
    the selected **top-K features**.

-   The pipeline computes the **average accuracy** on the test set.
    These accuracies are stored for later comparisons between
    classifiers allowing us to identify the value of `K` that gives the
    best performance.

```{r pipeline_structure}
# ------------------------------
# Function for pipeline workflow
# ------------------------------
run_pipeline <- function(X_train, X_test, y_train, y_test, topk = 10){
  
  # 1. Extract parameters for Z-score scaling
  mu = apply(X_train, 2, mean)
  sigma = apply(X_train, 2, sd)
  
  # 2. Scale train and test partitions with TRAIN parameters
  X_train_sc = scale(X_train, center = mu, scale = sigma)
  X_test_sc = scale(X_test, center = mu, scale = sigma)
  
  # 3. Select K Best
  topk = min(topk, ncol(X_train_sc)) # "topk" > ncol: not allowed 
  top_k = SelectKBest(X = X_train_sc, y = y_train, K = topk)
  
  # 4. Select K features and turn matrix into dataframe format
  X_train_df = data.frame(y = y_train, X_train_sc[,top_k, drop = FALSE])
  X_test_df = data.frame(X_test_sc[,top_k, drop = FALSE])
  
  # 5. LogReg classification
  acc_LogReg = LogReg_classifier(X_train_df, X_test_df, y_test)
  
  # 6. GNB classification
  acc_GNB = GNB_classifier(X_train_df, X_test_df, y_test)
  
  # 7. LDA classification
  acc_LDA = LDA_classifier(X_train_df, X_test_df, y_test)
  
  # 8. QDA classification
  acc_QDA = QDA_classifier(X_train_df, X_test_df, y_test)
  
  # Results
  return(list(LogReg=acc_LogReg,GNB=acc_GNB,LDA=acc_LDA,QDA=acc_QDA,features=top_k))
}
```

The previous pipeline is extended to **5-Fold Cross-Validation** to
obtain a more robust estimate of performance for each number of top
features (`topk`).

**Fold creation and initialization**:

-   The dataset is split into **5 folds**, and some variables are
    defined **to store accuracies** and the selected **feature subsets**
    for each classifier and fold.

**Training and testing per fold**

-   For each fold, one subset is used as the test set, while
    the remaining 4 folds form the training set.

-   This ensures that every data point is used once for testing and 4
    times for training. In addition, each classifier will be trained
    using the same fold assignments **to avoid data leakage** and to
    ensure **generalization**.

**Pipeline execution and data collection**

-   The previously `pipeline` (scaling, feature selection, classifier
    training) is applied to the training and testing sets in each fold.

-   **Accuracy scores** and the **selected top-K features** are recorded
    for each classifier.

-   After all folds, the **average performance** is computed to identify
    the optimal number of features and compare classifiers.

```{r kfold_cv}
# Structure for K-Fold CV Wrapper
kfold_cv <- function(X, y, k = 5, topk = 10){
  folds = createFolds(y, k = k, list = TRUE, returnTrain = TRUE)
  features_list = vector("list", k)
  # Initialize with NA so failed folds don't count as "0% accuracy"
  scores = matrix(NA, nrow = 4, ncol = k, 
                  dimnames = list(c("LogReg", "GNB", "LDA", "QDA"), NULL))

  for (i in seq_along(folds)){
      idx_train = folds[[i]]
      X_train_fold = X[idx_train,]
      y_train_fold = y[idx_train]
      X_test_fold = X[-idx_train,]
      y_test_fold = y[-idx_train]
      
      res = run_pipeline(X_train_fold, X_test_fold, y_train_fold, y_test_fold, topk)
      scores[, i] = c(res$LogReg, res$GNB, res$LDA, res$QDA)
      features_list[[i]] = res$features 
  }

  # STABILITY SELECTION to avoid selection bias
  # 1. Unlist all features chosen across the folds
  all_selected <- unlist(features_list)
  
  # 2. Count the frequency of each feature
  feature_counts <- table(all_selected)
  
  # 3. Sort them by most frequent to least frequent, and pick the top K
  sorted_features <- sort(feature_counts, decreasing = TRUE)
  final_features <- names(sorted_features)[1:topk]
  
  out = list(cv_scores = scores, 
             feat_LogReg = final_features,
             feat_GNB    = final_features,
             feat_LDA    = final_features,
             feat_QDA    = final_features)
  return(out)
}
```

This is an output example of the previous process.

```{r kfold_cv_example}
# kfold_cv(X = X_train, y = classes[idx_train], k = 5, topk = 5)

# Output:
# $cv_scores
#             [,1]      [,2]      [,3]      [,4]      [,5]
# LogReg 0.7800000 0.8145695 0.8211921 0.7894737 0.8026316
# GNB    0.7800000 0.7748344 0.7947020 0.7697368 0.7697368
# LDA    0.7800000 0.8211921 0.8211921 0.7960526 0.8026316
# QDA    0.8066667 0.8013245 0.8410596 0.8157895 0.7894737
# 
# $feat_LogReg
# [1] "APP_N"   "S6_N"    "pPKCG_N" "GluR3_N" "SYP_N"  
# 
# $feat_GNB
# [1] "APP_N"   "S6_N"    "pPKCG_N" "GluR3_N" "SYP_N"  
# 
# $feat_LDA
# [1] "APP_N"   "S6_N"    "GluR3_N" "pPKCG_N" "SYP_N"  
# 
# $feat_QDA
# [1] "APP_N"   "S6_N"    "pPKCG_N" "GluR3_N" "SYP_N"  
```

In the following chunk, the **full pipeline** will be executed for each
number of top features (`topk`). Performance metrics, such as
**accuracies** and **standard deviations**, will be computed, jointly
with the corresponding **selected feature subsets**.

QDA is sensitive to **high dimensionality** and has shown **numerical
stability issues** at the covariance matrix calculation step. Therefore,
we decided to build an array to **track** exactly when and where these
**failures** occur during cross-validation. As we iteratively increase
the number of features, we record the number of folds in which QDA model
fails to "converge". This information is then used to generate a
diagnostic plot, showing the dimensionality at which the model stability
breaks down.

```{r kfold_cv_pipeline}
set.seed(619) 

max_feats = min(30, length(proteins)) # Define the number of features

qda_stability <- numeric(max_feats) # Initialize vector to track qda failures

# Define the result containers (mean, sd, best_feats) for each classifier
mean_acc = matrix(0, nrow = 4, ncol = max_feats, 
                  dimnames = list(c("LogReg", "GNB", "LDA", "QDA"), NULL))

sd_acc = matrix(0, nrow = 4, ncol = max_feats,
                dimnames = list(c("LogReg","GNB","LDA","QDA"), NULL))

best_feats = list(LogReg = vector("list", max_feats), 
                  GNB = vector("list", max_feats),
                  LDA = vector("list", max_feats),
                  QDA = vector("list", max_feats))

# Run the K-Fold CV pipeline in a loop ---> 1:max_features
for (i in 1:max_feats){
  # 5-Fold CV
  kfold_result = kfold_cv(X = X_train, y = classes[idx_train], k = 5, topk = i)
  # Check the number of folds where QDA is failing
  qda_stability[i] <- sum(is.na(kfold_result$cv_scores["QDA", ]))
  # Save 
  mean_acc[, i] = rowMeans(kfold_result$cv_scores, na.rm = TRUE)
  sd_acc[, i]   = apply(kfold_result$cv_scores, 1, sd, na.rm = TRUE)
  
  best_feats$LogReg[[i]] = kfold_result$feat_LogReg
  best_feats$GNB[[i]]    = kfold_result$feat_GNB
  best_feats$LDA[[i]]    = kfold_result$feat_LDA
  best_feats$QDA[[i]]    = kfold_result$feat_QDA
}
```

We monitored the numerical stability of QDA, which exhibited significant
rank deficiency as the number of features increased. Specifically, at
feature counts above 15, QDA failed to converge in multiple folds. Our
suggestion for QDA is to select as 'Optimal K' approx. 10–13 features,
ensuring that we operate within the 'Stability Zone' of the QDA
algorithm.

```{r, echo = FALSE, fig.height=4.5, fig.width=6, fig.align="center"}
# ----------------------
# QDA stability checking
# ----------------------
first_fail <- which(qda_stability > 0)[1] # First failure for the vertical line

plot(1:max_feats, qda_stability, type = "s", col = "red", lwd = 2,
     xlab = "Number of Features (top-k)", ylab = "Folds Failed (out of 5)",
     main = "QDA Mathematical Stability Limit")
grid()

if(!is.na(first_fail)) {
  abline(v = first_fail - 0.5, col = "darkgreen", lty = 2)
  text(first_fail/2, 4.5, "Stable Zone", col = "darkgreen", font = 2)
  text(first_fail + (max_feats-first_fail)/2, 4.5, "Failure Zone", col = "red", font = 2)
}
```

The behavior of the classifiers for different feature subsets is shown
in graphs. Each classifier was trained using the same folds to avoid
data leakage. The mean and standard deviation across the 5-Folds are
computed as performance metrics for each feature subset.

Average accuracy is plotted against the number of selected features. We
use a 2x2 grid, with one graph per classifier, to visualize the general
performance from the first to the maximum number of features. The upper
and lower bounds represent the standard deviation across the 5-Folds,
allowing us to observe the performance variability.

To select the optimal number of features, we estimate the **elbow
point** as the point that maximizes the orthogonal distance to the
straight line connecting the first and last values of the curve. This
procedure is applied to the GNB, LDA, and LogReg classifiers to ensure
methodological consistency. However, for QDA, the preferred approach is
to select an intermediate value between 10 and 13 features, as discussed
previously.

```{r, echo = FALSE}
# The code is defined as the point which maximizes the orthogonal distance 
# to the straight line that connects the first and last value of the curve.
get_elbow_point <- function(m){
  n = length(m)
  x = 1:n
  # Starting point: (x1,y1)
  x1 = 1
  y1 = m[1]
  # Ending point: (x2, y2)
  x2 = n
  y2 = m[n]
  # Straight line coefficients
  a = y2-y1
  b = x1-x2
  c = x2*y1 - x1*y2
  # Orthogonal distance
  dist = abs(a*x + b*m + c) / sqrt(a^2 + b^2)
  return(which.max(dist))# elbow point position
}

plot_elbow_curve <- function(max_feats, m, s, main_title) {
  
  elbow_idx = get_elbow_point(m)# Get elbow point position
  
  y_min <- min(m - s, na.rm = TRUE) # axis-y lim
  y_max <- 1.0 # axis-y lim
  
  # Plot the Accuracy vs Top-K Features
  plot(1:max_feats, m, type = "b", pch = 19, col = "blue",
       xlab = "Number of top features", ylab = "Mean CV Accuracy",
       main = main_title, ylim = c(y_min - 0.05, y_max + 0.05))
  # Plot standard deviation bounds
  lines(1:max_feats, m + s, lty = 2, col = "gray40")
  lines(1:max_feats, m - s, lty = 2, col = "gray40")
  
  # Plot the straight line between first and last point
  segments(x0 = 1, y0 = m[1], x1 = max_feats, y1 = m[max_feats], 
           col = "darkgreen", lty = 2, lwd = 2)
  
  points(elbow_idx, m[elbow_idx], col = "purple", pch = 19, cex = 1.5)
  
  legend("bottomright", 
         legend = c(expression(mu), expression(mu %+-% sigma), "Secant", "Elbow"),
         col = c("blue", "gray40", "darkgreen", "purple"),
         lty = c(1, 2, 2, 1), 
         pch = c(19, NA, NA, 19), 
         bty = "n", cex = 0.95)
  grid()
}
```

```{r accuracy_classifiers, echo = FALSE, fig.width=10, fig.height=8, out.width="100%"}
# Minimum value
y_min = min(mean_acc[c("LogReg","GNB","LDA","QDA"), ], na.rm = TRUE)
# Margins for plots
par(mfrow = c(2,2), mar = c(4, 4, 2.5, 1))
# -------------------
# Logistic Regression
# -------------------
m = mean_acc["LogReg",] # average value
s = sd_acc["LogReg",] # standard deviation
plot_elbow_curve(max_feats, m, s, "LogReg: Accuracy vs Top-K Features")
# --------------------
# Gaussian Naive Bayes
# --------------------
m = mean_acc["GNB",] # average value
s = sd_acc["GNB",] # standard deviation
plot_elbow_curve(max_feats, m, s, "GNB: Accuracy vs Top-K Features")
# ----------------------------
# Linear Discriminant Analysis
# ----------------------------
m = mean_acc["LDA",] # average value
s = sd_acc["LDA",] # standard deviation
plot_elbow_curve(max_feats, m, s, "LDA: Accuracy vs Top-K Features")
# -------------------------------
# Quadratic Discriminant Analysis
# -------------------------------
m = mean_acc["QDA",] # average value
s = sd_acc["QDA",] # standard deviation
plot(1:max_feats, mean_acc["QDA",], type = "b", pch = 19, col = "blue",
     xlab = "Number of top features", ylab = "Mean CV Accuracy",
     main = "QDA: Accuracy vs Top-K Features", ylim=c(y_min-0.1, 1.0))
lines(1:max_feats, m + s, lty=2, col="gray40") # up sigma
lines(1:max_feats, m - s, lty=2, col="gray40") # down sigma
segments(x0 = 11, y0 = 0, x1 = 11, y1 = m[11], 
         col = "purple", lty = 2, lwd = 2)
points(11, m[11], col = "purple", pch = 19, cex = 1.5)
legend("bottomright", legend = c(expression(mu), expression(mu %+-% sigma),"Elbow"),
       col=c("blue","gray40","purple"),lty=c(1,2,3),pch=c(19, NA, 17),
       bty="n",cex=1.1)
grid()
```

**Observations**:

LogReg (`k=14`) and LDA (`k=12`) show similar behavior, with a positive
trend that stabilizes as the number of features increases. In contrast,
GNB (`k=15`) exhibits a nearly constant trend across all number of top
features, and displays the largest variability, as shown by the wider
standard deviation bands.

QDA (`k=11`) exhibits a positive trend with narrower standard deviation
bands, but as explained in the previous plot, it has stability issues
beyond 15 features due to the high dimensionality. After this point,
some folds fail, and the results become unreliable. We selected the
elbow point based on the recommendation previously discussed.

## Model Training

Based on the trends observed in the accuracy curves and the variability
of each classifier, we determined an appropriate number of top features
for training. For each classifier, the 'Optimal K' was chosen based on
the elbow point of the cross-validation accuracies, while considering
the stability limitations of QDA too. Following Occam's razor, we will
prefer simpler models to improve interpretability and computational
efficiency, while avoiding issues such as rank deficiency in covariance
matrices.

```{r scale_training}
# Extract standardization parameters and scale data
mu = colMeans(X_train)
sigma = apply(X_train, 2 , sd)
X_train_sc = scale(X_train, center = mu, scale = sigma)
X_test_sc = scale(X_test, center = mu, scale = sigma)
# Define labels for testing set
y_test = classes[-idx_train]
```

```{r logreg}
optimal_k = 14
# 1. Logistic Regression
best_k_LogReg = best_feats$LogReg[[optimal_k]]

# 1.1. Build dataframes
df_train_LogReg = data.frame(X_train_sc[, best_k_LogReg],class = classes[idx_train])
df_test_LogReg  = data.frame(X_test_sc[, best_k_LogReg])

# 1.2. Build model and predict df_test
logreg_model = glm(class ~ ., data = df_train_LogReg, 
                   family = binomial(link = "logit"))
logreg_probs = predict(logreg_model, newdata = df_test_LogReg, type = "response")
logreg_class = factor(ifelse(logreg_probs > 0.5, "Ts65Dn", "Control"),levels = lvls)
```

```{r GNB}
optimal_k = 15
# 2. Gaussian Naive Bayes
best_k_GNB = best_feats$GNB[[optimal_k]]

# 2.1. Build dataframes
df_train_GNB = data.frame(X_train_sc[, best_k_GNB], class = classes[idx_train])
df_test_GNB  = data.frame(X_test_sc[, best_k_GNB])

# 2.2. Build model and predict df_test
gnb_model = naiveBayes(class ~ ., data = df_train_GNB)
gnb_probs = predict(gnb_model, df_test_GNB, type = "raw")
gnb_class = predict(gnb_model, df_test_GNB, type = "class")
```

```{r LDA}
optimal_k = 12
# 3. LDA
best_k_LDA = best_feats$LDA[[optimal_k]]

# 3.1. Build dataframes
df_train_LDA = data.frame(X_train_sc[, best_k_LDA], class = classes[idx_train])
df_test_LDA  = data.frame(X_test_sc[, best_k_LDA])

# 3.2. Build model and predict df_test
lda_model = lda(class ~ ., data = df_train_LDA)
lda_pred = predict(lda_model, df_test_LDA)
lda_probs = lda_pred$posterior
lda_class = lda_pred$class
```

```{r QDA}
optimal_k = 11
# 4. QDA
best_k_QDA = best_feats$QDA[[optimal_k]]

# 4.1. Build dataframes
df_train_QDA = data.frame(X_train_sc[, best_k_QDA], class = classes[idx_train])
df_test_QDA  = data.frame(X_test_sc[, best_k_QDA])

# 4.2. Build model and predict df_test
qda_model = qda(class ~ ., data = df_train_QDA)
qda_pred = predict(qda_model, df_test_QDA)
qda_probs = qda_pred$posterior
qda_class = qda_pred$class
```

# Results & Evaluation

## Robust Cross-Validation

Single train/test splits can be noisy. To get a robust estimate of
performance, we perform **Stratified 10-Fold Cross-Validation** on all
four models.

```{r robust cross validation, fig.width=10, fig.height=5, out.width="100%"}
set.seed(619);
# Define Control
train_control <- trainControl(method = "cv", number = 10, 
                              classProbs = TRUE, summaryFunction = twoClassSummary)
# Train Models using Caret
cv_logreg <- train(class ~ ., data = df_train_LogReg, method = "glm", 
                   family = "binomial", metric = "ROC", trControl = train_control)

cv_lda <- train(class ~ ., data = df_train_LDA, method = "lda", 
                metric = "ROC", trControl = train_control)

cv_qda <- train(class ~ ., data = df_train_QDA, method = "qda", 
                metric = "ROC", trControl = train_control)

cv_nb <- train(class ~ ., data = df_train_GNB, method = "nb", 
               metric = "ROC", trControl = train_control)

# Compile & Visualize
cv_results <- resamples(list(LogReg = cv_logreg, 
                             LDA = cv_lda, 
                             QDA = cv_qda, 
                             NaiveBayes = cv_nb))

dotplot(cv_results, main = "10-Fold CV Performance (ROC/AUC)")
```

**Observation**: GNB exhibits the worst performance compared to the
other models. This is expected because GNB assumes that the expression
level of each protein is independent of the others. In a real biological
system, this assumption is not entirely accurate, as proteins interact
within pathways and networks. The results indicate that the relationship
between features are more complex than the model assumes.

## Model Comparison (Test Set)

In our analysis, the target variable was encoded such that trisomic
genotype (Ts65Dn) represents the positive class (`lvl=1`) and the normal
genotype (Control) represents the negative class (`lvl=2`). Therefore,
in this scenario, **sensitivity** measures the ability of the model to
correctly classify Ts65Dn, while **specificity** measures its accuracy
in identifying Control.

We compare the models based on Accuracy, Sensitivity (Recall), and
Specificity on the hold-out test set.

```{r model comparison, echo = FALSE}
LogReg_cm = confusionMatrix(logreg_class, y_test, positive = "Ts65Dn")
GNB_cm = confusionMatrix(gnb_class, y_test, positive = "Ts65Dn")
LDA_cm = confusionMatrix(lda_class, y_test, positive = "Ts65Dn")
QDA_cm = confusionMatrix(qda_class, y_test, positive = "Ts65Dn")

comparison <- data.frame(
  Method = c("Logistic Regression", "LDA", "QDA", "Naive Bayes"),
  Accuracy = c(LogReg_cm$overall["Accuracy"], 
               LDA_cm$overall["Accuracy"], 
               QDA_cm$overall["Accuracy"], 
               GNB_cm$overall["Accuracy"]),
  Sensitivity = c(LogReg_cm$byClass["Sensitivity"], 
                  LDA_cm$byClass["Sensitivity"], 
                  QDA_cm$byClass["Sensitivity"], 
                  GNB_cm$byClass["Sensitivity"]),
  Specificity = c(LogReg_cm$byClass["Specificity"], 
                  LDA_cm$byClass["Specificity"], 
                  QDA_cm$byClass["Specificity"], 
                  GNB_cm$byClass["Specificity"])
)
```

```{r visualize comparison, echo = FALSE, fig.height=3.25, fig.width=6}
comparison_long <- comparison %>% pivot_longer(cols = -Method, 
                                               names_to = "Metric", 
                                               values_to = "Value")

ggplot(comparison_long, aes(x = Method, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Performance Comparison of Probabilistic Classifiers", 
       y = "Value", x = "") +
  theme_minimal() + 
  scale_fill_brewer(palette = "Set2") + 
  coord_flip() + 
  theme(legend.position = "bottom")
```

```{r, echo = FALSE}
knitr::kable(comparison, digits = 4, caption = "Performance Comparison")
```

**Interpretation:**

When observing the results from the probabilistic and discriminative
classifiers on the expression levels of the selected proteins, the
models LogReg, LDA, and QDA each offer **different advantages**
depending on the specific context of the study.

In terms of **overall predictive power**, **LDA** achieves the highest
accuracy (0.9475) and specificity (0.9649) in the analysis. Its
**maximum specificity** makes it the most reliable model for correctly
identifying subjects with normal genotypes (Control). In contrast,
**LogReg** demonstrates the most **balanced performance** across metrics
and achieves the **highest sensitivity** (0.9346), becoming particularly
effective at detecting the perturbed protein pathways of the Ts65Dn
group.

Finally, **QDA** achieves a highly competitive accuracy (0.9383), and
demonstrates sensitivity (0.9281) close to that of LDA. However, this
model is strongly affected by **multicollinearity issues**, making it
less suitable for high-dimensional data. While QDA could be reasonable
when working with a small subset of biomarkers, in this context its
**mathematical complexity** becomes a disadvantage.

## ROC Curves

To evaluate the discriminatory capability independent of a specific
threshold, we analyze the Receiver Operating Characteristic (ROC)
curves. The **LogReg** (blue), **LDA** (red), and **QDA** (purple) show
the steepest rise toward the top-left corner, confirming their high
performance in terms of AUC and overall discriminatory power.

```{r ROC, echo = FALSE, fig.height=3.25, fig.width=6, width.out = "100%", fig.align="center"}

roc_logreg = roc(y_test, logreg_probs, levels = lvls, direction = "<")
roc_lda = roc(y_test, lda_probs[, "Ts65Dn"], levels = lvls, direction = "<")
roc_qda = roc(y_test, qda_probs[, "Ts65Dn"], levels = lvls, direction = "<")
roc_gnb = roc(y_test, gnb_probs[, "Ts65Dn"], levels = lvls, direction = "<")

roc_data <- bind_rows(data.frame(Sensitivity = roc_logreg$sensitivities, 
                                 Specificity = roc_logreg$specificities, 
                                 Method = "Logistic", 
                                 AUC = as.numeric(roc_logreg$auc)),
                      data.frame(Sensitivity = roc_lda$sensitivities, 
                                 Specificity = roc_lda$specificities, 
                                 Method = "LDA", 
                                 AUC = as.numeric(roc_lda$auc)),
                      data.frame(Sensitivity = roc_qda$sensitivities, 
                                 Specificity = roc_qda$specificities, 
                                 Method = "QDA", 
                                 AUC = as.numeric(roc_qda$auc)),
                      data.frame(Sensitivity = roc_gnb$sensitivities, 
                                 Specificity = roc_gnb$specificities, 
                                 Method = "Naive Bayes", 
                                 AUC = as.numeric(roc_gnb$auc))
)

ggplot(roc_data, aes(x = 1 - Specificity, y = Sensitivity, color = Method)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(title = "ROC Curves", x = "False Positive Rate (1 - Specificity)", y = "Sensitivity") +
  theme_minimal() + scale_color_brewer(palette = "Set1") + coord_equal()
```

```{r AUC}
# Compute AUC for each model
auc_table <- data.frame(
  Method = c("Logistic Regression", "LDA", "QDA", "Naive Bayes"),
  AUC = c(
    as.numeric(auc(roc_logreg)),
    as.numeric(auc(roc_lda)),
    as.numeric(auc(roc_qda)),
    as.numeric(auc(roc_gnb))
  )
)

# Merge with existing comparison table
comparison <- comparison %>%
  left_join(auc_table, by = "Method") %>%
  relocate(AUC, .after = Accuracy)

knitr::kable(comparison, digits = 4, 
             caption = "Performance Comparison (Test Set)")
```

## Threshold Exploration (Medical Context Tuning)

By default, probabilistic classifiers such as LDA and LogReg use a 0.5
decision threshold to assign class labels, assuming that False Positives
and False Negatives have similar cost. However, in medical diagnostics,
this default threshold is rarely optimal, as the cost of a False
Negative (failing to identify a trisomic mouse) may differ significantly
from a False Positive. To address this, we built a plot to study the
effect of tuning the decision threshold on the Accuracy, Sensitivity,
and Specificity. This visualization allows us to assess the trade-off
between metrics beyond the default 0.5 cutoff. This tool becomes really
useful depending on the specific requirements of the analysis.

```{r, echo = FALSE}

lvls <- c("Ts65Dn", "Control") # Define levels 
thresholds <- seq(0.05, 0.95, by = 0.05) # Define threshold grid

calculate_metrics <- function(probs, actual_labels) {
  map_dfr(thresholds, function(t) {
    
    pred <- factor(ifelse(probs > t, "Ts65Dn", "Control"), levels = lvls)
    cm <- confusionMatrix(pred, actual_labels, positive = "Ts65Dn")
    data.frame(Threshold = t,
               Accuracy = cm$overall["Accuracy"],
               Sensitivity = cm$byClass["Sensitivity"],
               Specificity = cm$byClass["Specificity"])
  })
}

prob_lda = lda_probs[, "Ts65Dn"] 
prob_logreg = logreg_probs

res_lda = calculate_metrics(prob_lda, y_test)
res_logreg = calculate_metrics(prob_logreg, y_test)
```

```{r, echo = FALSE, fig.height=4.25, fig.width=6.5, width.out = "100%", fig.align="center"}
# Plot threshold effects
plot_thresholds <- function(results_df, model_title) {
  threshold_long <- results_df %>%
    pivot_longer(cols = -Threshold, names_to = "Metric", values_to = "Value") %>%
    drop_na()
  
  ggplot(threshold_long, aes(x = Threshold, y = Value, color = Metric)) +
    geom_line(linewidth = 1) +
    geom_point() +
    geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
    labs(title = model_title,
         x = "Decision Threshold", 
         y = "Metric Value") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
    scale_color_brewer(palette = "Set2")
}

# Individual plots
plot_lda = plot_thresholds(res_lda, "LDA")
plot_logreg = plot_thresholds(res_logreg, "LogReg")

# Concatenate both plots
final_plot = plot_lda + plot_logreg + 
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Effect of Decision Threshold",
    subtitle = "Dashed line shows the default 0.5 threshold",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
                  plot.subtitle = element_text(hjust = 0.5, size = 10))
  ) & 
  theme(legend.position = "bottom", legend.title = element_blank())

final_plot
```
